Understanding the nanoscale structure of hexagonal phase lyotropic liquid crystal membranes

### 2018-04-24 Ben_Manuscripts/structure_paper/Draft.tex
removed unneeded files
3D fourier transforms of simple systems
Edits and new figures

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{subcaption}
\usepackage{amsmath} 
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{gensymb}

%MRS5: could be punchier. I will keep thinking.  I should be the last author.  Chen the names that Joe and Matt wanttu use (presumably with some middle initial?
\title{Understanding the nanoscale structure of hexagonal phase lyotropic 
liquid crystal membranes}
\author{Benjamin J. Coscia}
\author{Richard D. Noble}
\author{Douglas L. Gin}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}
\alsoaffiliation{Department of Chemistry and Biochemistry, University of Colorado Boulder, Boulder CO 80309, USA}
\author{Joe Yelk}
\author{Matthew Glaser}
\affiliation{Department of Physics, University of Colorado Boulder, Boulder CO, 80309, USA}
\author{Xunda Feng}
\affiliation{Department of Chemical and Environmental Engineering, Yale University, New Haven, Connecticut 06511, USA}
\author{Michael R. Shirts}
\email{michael.shirts@colorado.edu}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}
\begin{document}

@@

@@

   %MRS4: these two paragraphs are perhaps not ideal.  They describe the process
  %of making the membranes, but that is not what we need as the introduction to
  %THIS paper.  Instead, you need to focus on the strength/issues with RO and
  %NF that motivate this research.
  % BJC3: The point I am trying to make is that the way the membranes are made 
  % results in either low selectivity or low permeability. I could just state
  % that, or give a little background on the processes to make the point more
  % tangible

@@

%MRS6: formatting suggestion
  \item \textit{If layers do exist, how many monomers constitute a single layer? \label{point:monomernum}} 
  A simple molecular simulation study of a similar molecule suggested that
  there are 4 monomers in each layer \cite{zhu_methacrylated_2006}. Their model
  likely does not sufficiently imitate the chemical environment present in the
  real system since it only contains 16 total monomers. A separate calculation
  for NA-GA3C11, based on the estimated volume of the liquid crystal monomers,
  proposes that each layer consists of seven
  monomers\cite{resel_structural_2000}. Our best chance to answer this question
  is by using a molecular model orders of magnitude larger than any other
  reported atomistic liquid crystal membrane simulation, as we present here. We
  will directly change the layer composition and note its effect on membrane
  structure.

  \item \textit{How do monomers in each layer position themselves with respect to
  surrounding layers? \label{point:orientation}}
  $\pi$-$\pi$ stacking interactions between aromatic monomer head groups may be
  a driving force of self-assembly in this system \cite{gazit_possible_2002}. Gas
  phase \textit{ab initio} studies of benzene dimers have shown a clear energetic
  advantage for parallel displaced and T-shaped $\pi$-$\pi$ stacking
  conformations over a sandwiched conformation \cite{sinnokrot_estimates_2002}.
  Substituted benzene rings exhibit an even stronger $\pi$-$\pi$ stacking
  attraction which favors the parallel displaced configuration in all cases
  except where the substitutions are extremely electron
  withdrawing\cite{waller_hybrid_2006,ringer_effect_2006}. In this study, we
  compare simulated X-ray diffraction patterns to experiment in order to deduce
  which stacking configuration is most likely. We also dive deeper into the
  origin of features present in the 2D diffraction patterns.

  \item \textit{Does our model support the existence of layers and if so, how well
  defined are the layers? \label{point:layers}} 
  %BJC: Reordered, so changing the wording here.
  Indications of strong $\pi$-$\pi$ stacking interactions in the direction
  perpendicular to the membrane plane, combined with the wedge-like character of
  the LLC monomers, suggests that monomers should arrange into disks that stack in
  layers. $\pi$-$\pi$ stacking will only occur between the aromatic
  monomer head groups which leaves no description of what is happening in the
  monomer tail region. The tails may entangle isotropically while maintaining
  stacking order among headgroups.

  \item \textit{Can the system exist in other metastable states or phases that are not
  accessed during experiments? \label{point:metastable}}
  There remains the possibility that there is more than one metastable state
  associated with a given LLC phase. Simulating a membrane atomistically
  requires many atoms which limits the timescales acessible with MD. It is
  reasonable to expect that we will generate configurations which are kinetically
  trapped in a metastable free energy basin. We must be able to identify which
  states 
  %MRS6: best not to set up too high expectations
  %is produced experimentally.
  are most consistent with experiment.

  \item \textit{What constitutes a pore, and what are the details of its pore architecture? \label{point:poredefinition}}
  The limited picture that experiment provides tells us that there are
  hexagonally packed, hydrophilic regions where transport is likely to occur.
  One may instinctively assume that these regions are hollow tube-like pathways.
  We will explore the composition of the pores and the partition between the
  hydrophilic and hydrophobic regions.

@@

  \subsubsection{Number of monomer per layer}

  Our simulations best support a model built with 5 monomers per layer based on
  the measured equilibrated pore-to-pore distances. To discern the composition of
  the monomer layers, addressing question (\ref{point:monomernum}), we ran
  simulations of systems created with 4--8 monomers per layer. We built systems
  in both the parallel displaced and sandwiched configurations with layers
  initially spaced 3.7 \AA~apart. We prepared equilibrated configurations
  according to the dry equilibration procedure. All systems are stable after 400
  ns of simulation. 
  %MRS6: Above, probably need to be a bit more clear what you mean by stable (no noticable drift in pore-to-pore), and if you are only using data after the point of stability.
  %BJC2: I believe this is included in methods. I can restate if that's not clear
In a sense, all systems are at least metastable, addressing
  question (\ref{point:metastable}), however not all will make physical sense or fit
  the experimental profile that we are trying to match. Figure~\ref{fig:p2p}
  shows 
%MRS6:
%the pore spacing 
the equilibrated pore-to-pore distances 
for all systems tested. 
Systems built with 5 monomers in
  each layer equilibrate to a pore spacing that is most consistent with the
  experimental value of 4.12 nm derived from SAXS measurements
  (Figure~\ref{fig:SAXS}). 
%MRS6: discuss very briefly how far off 4 and 6 are to motivate selecting 5 exclusively.
The remainder of this discussion will focus on the
  analysis of systems built with 5 monomers per layer.

  \begin{figure}
	\centering
	\includegraphics[width=\linewidth]{p2p.png}
        %MRS6: should there be a grapical demonstration of sandwiched vs. parallel displaced, especially equilbrated?
        %MRS6: some tweaks below.
	\caption{Systems with 5 monomers per layer have
		equilibrated pore spacing closest to the experimental value of 4.12 nm. The
			equilibrated pore spacing of the model increases as the number of monomers in
			each layer increases. The pore spacing of systems starting in the sandwiched
			configuration is systematically lower than those started in the parallel
			displaced configuration.}~\label{fig:p2p}
  \end{figure}

  \subsubsection{Relative interlayer head group orientation}

  We answer question (\ref{point:orientation}) by simulating X-ray diffraction
  patterns produced from equilibrated MD trajectories. We tested systems built
  with 5 monomers per layer in the parallel displaced and sandwiched
  configurations. We generated simulated patterns using portions of simulation
  trajectory after equilibration. The patterns for both structures are shown and
  compared to experiment in Figure \ref{fig:XRDsim}.

@@

  The parallel displaced configuration results in a simulated XRD pattern with
  the closest match to experiment. It produces the only pattern that exhibits all
  major reflections. R-alkanes and R-pores appear as they do in the sandwiched
  configuration. R-spots and R-$\pi$ appear, however with a lower intensity
  relative to R-alkanes when compared to the sandwiched configuration. R-helix
  appears due to the parallel displaced aromatic rings. It is a subharmonic of
  R-$\pi$ since the nearest vertically stacked head group to any given head group
  is $2\times$R-$\pi$~away. 
  %BJC: adjust figure size and alignment -- probably easiest to set figure size in matplotlib
  \begin{figure}
  \begin{subfigure}{0.3\linewidth}
        \centering
        \vspace{-0.2em}
        \includegraphics[width=1.1\linewidth,trim={1cm 0 1.3cm 0},clip]{offset_rzplot.png}
        \caption{}~\label{fig:rz_offset}
  \end{subfigure}
  \begin{subfigure}{0.3\linewidth}
        \centering
%        \vspace{0.25em}
        \includegraphics[scale=0.285]{WAXS_raw.png}
        \caption{}~\label{fig:raw_waxs}
  \end{subfigure}
  \begin{subfigure}{0.3\linewidth}
        \centering
        \includegraphics[width=1.1\linewidth,trim={1cm 0 1.3cm 0},clip]{layered_rzplot.png}
        \caption{}~\label{fig:rz_layered}
  \end{subfigure}
  \begin{subfigure}{0.0544\linewidth}
        \centering
        \vspace{-4.00em}
        \includegraphics[width=\linewidth]{colorbar_seismic.png}
  \end{subfigure}
%MRS6: question: the SAXS region looks different between sandwiched and parallel displaced. Any thoughts why? We can discuss in person.  And seems like have to comment on the difference in intensity between Sandwich and parallel displaced in the pi-pi region.
%MRS6: I wonder to what extent showing the reflections of tails removed would make it clearer pi-pi is really there.  Maybe in supporting info?
%BJC2: Sure, I'll add that to supporting info
  \caption{The simulated XRD pattern generated from the equilibrated trajectory
	  of the parallel displaced configuration (a) exhibits all major reflections
	  present in the experimental WAXS pattern (b). The simulated XRD pattern
	  generated from the equilibrated trajectory in the sandwiched configuration (c) shows
	  all major reflections except R-helix.}
  \label{fig:XRDsim}
  \end{figure}

@@

  The R-spots signal, which appears in both simulated XRD patterns, is not a
  result of alkane chain tilt. Previous literature has attributed the spots in
  this particular WAXS pattern as the product of tilted alkane
  chains~\cite{feng_scalable_2014}. We measured the tilt angle of the alkane
  chains and showed that our system equilibrates to an average tilt angle close
  to zero degrees (Fig.~\ref{fig:tilt}). To understand the origin of R-spots,
  we determined which atoms gave rise to the feature. Since R-spots is present as
  higher intensity spots within R-alkanes, it is likely that the spots arise as a
  consequence of the tails. By removing all non-tail atoms from the trajectory
  and simulating a diffraction pattern, we were able to isolate the cause of the
  spots to the tails (Figure~\ref{fig:tails}). Since the tails stay nearly flat,
  we plotted the centroids of the tails and measured the angle between each
  centroid and its nearest neighbors with respect to the plane of the membrane.
  We see distinct peaks in the distribution of these angles
  (Figure~\ref{fig:tail_packing}).

>>>>

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{subcaption}
\usepackage{amsmath} 
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{gensymb}
\usepackage{xr}

\makeatletter
\newcommand*{\addFileDependency}[1]{% argument=file name and extension
  \typeout{(#1)}
  \@addtofilelist{#1}
  \IfFileExists{#1}{}{\typeout{No file #1.}}
}
\makeatother

\newcommand*{\myexternaldocument}[1]{%
    \externaldocument{#1}%
    \addFileDependency{#1.tex}%
    \addFileDependency{#1.aux}%
}

\externaldocument[S-]{supplemental}

%MRS5: could be punchier. I will keep thinking.  I should be the last author.  Chen the names that Joe and Matt wanttu use (presumably with some middle initial?
\title{Understanding the nanoscale structure of hexagonal phase lyotropic 
liquid crystal membranes}
\author{Benjamin J. Coscia}
\author{Richard D. Noble}
\author{Douglas L. Gin}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}
\alsoaffiliation{Department of Chemistry and Biochemistry, University of Colorado Boulder, Boulder CO 80309, USA}
\author{Joe Yelk}
\author{Matthew A. Glaser}
\affiliation{Department of Physics, University of Colorado Boulder, Boulder CO, 80309, USA}
\author{Xunda Feng}
\affiliation{Department of Chemical and Environmental Engineering, Yale University, New Haven, Connecticut 06511, USA}
\author{Michael R. Shirts}
\email{michael.shirts@colorado.edu}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}
\begin{document}

@@

  Figure S~\ref{S-fig:monomer_color_coded}

@@

   %MRS4: these two paragraphs are perhaps not ideal.  They describe the process
  %of making the membranes, but that is not what we need as the introduction to
  %THIS paper.  Instead, you need to focus on the strength/issues with RO and
  %NF that motivate this research.
  % BJC2: The point I am trying to make is that the way the membranes are made 
  % results in either low selectivity or low permeability. I could just state
  % that, or give a little background on the processes to make the point more
  % tangible

@@

%BJC3: reworded a lot of these to be more in line with Matt's way of thinking about LC systems
%MRS6: formatting suggestion
%  \item \textit{If layers do exist, how many monomers constitute a single layer? \label{point:monomernum}} 
  \item \textit{How many monomers pack around a single pore column?\label{point:monomernum}} 
  A simple molecular simulation study of a similar molecule suggested that
  4 monomers radially pack around hydrophilic pore columns \cite{zhu_methacrylated_2006}.
  Their model likely does not sufficiently imitate the chemical environment present
  in the real system since it only contains 16 total monomers. A separate calculation
  for NA-GA3C11, based on the estimated volume of the liquid crystal monomers,
  proposes that each pore is surrounded by seven
  monomers\cite{resel_structural_2000}. Our best chance to answer this question
  is by using a molecular model orders of magnitude larger than any other
  reported atomistic liquid crystal membrane simulation, as we present here. We
  will directly change the density of monomers in each pore and note its effect
  on membrane structure.

%  \item \textit{How do monomers in each layer position themselves with respect to
%  surrounding layers? \label{point:orientation}}
  \item \textit{In what orientation do monomer head groups stack on top of each
  other \label{point:orientation} 
  $\pi$-$\pi$ stacking interactions between aromatic monomer head groups may be
  a driving force of self-assembly in this system \cite{gazit_possible_2002}. Gas
  phase \textit{ab initio} studies of benzene dimers have shown a clear energetic
  advantage for parallel displaced and T-shaped $\pi$-$\pi$ stacking
  conformations over a sandwiched conformation \cite{sinnokrot_estimates_2002}.
  Substituted benzene rings exhibit an even stronger $\pi$-$\pi$ stacking
  attraction which favors the parallel displaced configuration in all cases except
  where the substitutions are extremely electron withdrawing  
  \cite{waller_hybrid_2006,ringer_effect_2006}. We did not consider the T-shaped 
  configuration due to its geometric incompatibility with a hexagonal columnar phase.
  The entropic contribution of the long alkane tails should overwhelm any attraction
  between aromatic head groups in the T orientation. We limit our studies to
  comparison of systems built in the parallel displaced and sandwiched configurations.

% BJC3: I don't want to limit X-ray diffraction patterns just to deciding between
% parallel displaced and sandwiched. Since it doesn't actually decide either way.  
%  In this study, we
%  compare simulated X-ray diffraction patterns to experiment in order to deduce
%  which stacking configuration is most likely. We also dive deeper into the
%  origin of features present in the 2D diffraction patterns.

%  \item \textit{Does our model support the existence of layers and if so, how well
%  defined are the layers? \label{point:layers}}
%  Indications of strong $\pi$-$\pi$ stacking interactions in the direction
%  perpendicular to the membrane plane, combined with the wedge-like character of
%  the LLC monomers, suggests that monomers should arrange into disks that stack in
%  layers. $\pi$-$\pi$ stacking will only occur between the aromatic
%  monomer head groups which leaves no description of what is happening in the
%  monomer tail region. The tails may entangle isotropically while maintaining
%  stacking order among headgroups.  
% BJC3:rephrasing everything about the above question. 
  \item \textit{How strong are stacking correlations between monomer head groups?
  \label{point:layers}}

%BJC3: Going to talk about temperature rather than ordered/disordered states
%  \item \textit{Can the system exist in other metastable states or phases that are not
%  accessed during experiments? \label{point:metastable}}
%  There remains the possibility that there is more than one metastable state
%  associated with a given LLC phase. Simulating a membrane atomistically
%  requires many atoms which limits the timescales acessible with MD. It is
%  reasonable to expect that we will generate configurations which are kinetically
%  trapped in a metastable free energy basin. We must be able to identify which
%  states 
  %MRS6: best not to set up too high expectations
  %is produced experimentally.
%  are most consistent with experiment.

  \item \textit{What constitutes a pore, and what are the details of its composition?
  \label{point:poredefinition}}
  The limited picture that experiment provides tells us that there are
  hexagonally packed, hydrophilic regions where transport is likely to occur.
  One may instinctively assume that these regions are hollow tube-like pathways.
  We will explore the composition of the pores and the partition between the
  hydrophilic and hydrophobic regions.

@@

  \subsubsection{Density of monomers around pores}

  Our simulations best support a model built with 5 monomers per layer based on
  the measured equilibrated pore-to-pore distances. To discern the composition of
  the monomer layers, addressing question (\ref{point:monomernum}), we ran
  simulations of systems created with 4--8 monomers per layer. We built systems
  in both the parallel displaced and sandwiched configurations with layers
  initially spaced 3.7 \AA~apart. We prepared equilibrated configurations
  according to the dry equilibration procedure. The pore-to-pore spacing is stable
  in all systems are stable after 400 ns of simulation. Figure~\ref{fig:p2p} shows
  the equilibrated pore-to-pore distances for all systems tested. Systems built 
  with 5 monomers in each layer equilibrate to a pore spacing that is most consistent
  with the experimental value of 4.12 nm derived from SAXS measurements 
  (Figure~\ref{fig:SAXS}).

  The remainder of this discussion will focus on the analysis of systems built with
  5 monomers per layer. Systems built with 6 monomers per layer have an equilibrated
  pore spacing c.a. 0.50 nm higher than experiment. Systems built with 4 monomers per 
  layer equilibrate to a pore spacing 0.25 nm lower than experiment. In a sense, all
  systems tested are at least metastable, however not all will make physical
  sense or fit the experimental profile that we are trying to match. In the limit of
  infinite time, all simulations will converge to a single density. It is our 
  intention to choose an initial configuration which quickly reaches a monomer density
  close to experiment.

  \begin{figure}
	\centering
	\includegraphics[width=\linewidth]{p2p.png}
        %MRS6: should there be a grapical demonstration of sandwiched vs. parallel displaced, especially equilbrated?
        % BJC3: I have that in the supplemental information. I could move it into the main
        % body. I think it would be best in the intro where I pose the question of parallel 
        % displaced v. sandwiched.
	\caption{Systems with 5 monomers per layer have equilibrated pore spacings closest to
			 the experimental value of 4.12 nm. The equilibrated pore spacing of the model 	
			 increases as the number of monomers in each layer increases. The pore spacing
			 of systems starting in the sandwiched configuration are systematically lower
			 than those started in the parallel displaced configuration.}~\label{fig:p2p}
  \end{figure}

%  \subsubsection{Relative interlayer head group orientation}
  \subsubsection{Structural refinement}

  We further refined our structural understanding of the system by simulating X-ray 
  diffraction patterns produced from equilibrated MD trajectories and comparing them
  to experiment. We tested systems built with 5 monomers per layer in the parallel 
  displaced and sandwiched configurations at 300K. We generated simulated patterns using 
  portions of simulation trajectory after equilibration. The patterns for both 
  structures are shown and compared to experiment in Figure~\ref{fig:XRDsim}.

  Simulated XRD of the sandwiched configuration shows most all experimental features 
  except R-helix and R-spots. R-pores and R-alkanes appear in their expected locations. 
  R-pores is more intense than experiment, likely because we are simulating a perfect,
  infinite hexagonal array. The real system has defects and domain misalignment
  which decreases the overall intensity. R-$\pi$ appears with very high intensity and
  is shifted down to a lower value of $q_z$ meaning monomers in our model prefer to 
  stack farther apart.

@@

  Like the sandwiched configuration, the parallel displaced configuration creates
  a simulated XRD pattern with all major experimental reflections except for R-helix.
  R-alkanes and R-pores appear as they do in the sandwiched configuration. R-spots
  and R-$\pi$ appear, however with a lower intensity
  relative to R-alkanes when compared to the sandwiched configuration. R-helix
  appears due to the parallel displaced aromatic rings. It is a subharmonic of
  R-$\pi$ since the nearest vertically stacked head group to any given head group
  is $2\times$R-$\pi$~away. 
  %BJC: adjust figure size and alignment -- probably easiest to set figure size in matplotlib
  \begin{figure}
  \begin{subfigure}{0.3\linewidth}
        \centering
        \vspace{-0.2em}
        \includegraphics[width=1.1\linewidth,trim={1cm 0 1.3cm 0},clip]{offset_rzplot.png}
        \caption{}~\label{fig:rz_offset}
  \end{subfigure}
  \begin{subfigure}{0.3\linewidth}
        \centering
%        \vspace{0.25em}
        \includegraphics[scale=0.285]{WAXS_raw.png}
        \caption{}~\label{fig:raw_waxs}
  \end{subfigure}
  \begin{subfigure}{0.3\linewidth}
        \centering
        \includegraphics[width=1.1\linewidth,trim={1cm 0 1.3cm 0},clip]{layered_rzplot.png}
        \caption{}~\label{fig:rz_layered}
  \end{subfigure}
  \begin{subfigure}{0.0544\linewidth}
        \centering
        \vspace{-4.00em}
        \includegraphics[width=\linewidth]{colorbar_seismic.png}
  \end{subfigure}
%MRS6: question: the SAXS region looks different between sandwiched and parallel displaced. Any thoughts why? We can discuss in person.  And seems like have to comment on the difference in intensity between Sandwich and parallel displaced in the pi-pi region.
%MRS6: I wonder to what extent showing the reflections of tails removed would make it clearer pi-pi is really there.  Maybe in supporting info?
%BJC2: Sure, I'll add that to supporting info
  \caption{The simulated XRD pattern generated from the equilibrated trajectory
	  of the parallel displaced configuration (a) exhibits all major reflections
	  present in the experimental WAXS pattern (b). The simulated XRD pattern
	  generated from the equilibrated trajectory in the sandwiched configuration (c) shows
	  all major reflections except R-helix.}
  \label{fig:XRDsim}
  \end{figure}

@@

  The R-spots signal, which appears in both simulated XRD patterns, is not a
  result of alkane chain tilt. Previous literature has attributed the spots in
  this particular WAXS pattern as the product of tilted alkane
  chains~\cite{feng_scalable_2014}. We looked closer at the sandwiched
  configuration since R-spots appears with the most strength in its simulated XRD
  pattern. We measured the tilt angle of the alkane chains and showed that our
  system equilibrates to an average tilt angle close to zero degrees
  (Fig.~\ref{fig:tilt}). To understand the origin of R-spots, we determined which
  atoms gave rise to the feature. Since R-spots is present as higher intensity
  spots within R-alkanes, it is likely that the spots arise as a consequence of
  the tails. By removing all non-tail atoms from the trajectory and simulating a
  diffraction pattern, we were able to isolate the cause of the spots to the
  tails (Figure~\ref{fig:tails}). Since the tails stay nearly flat, we plotted
  the centroids of the tails and measured the angle between each centroid and its
  nearest neighbors with respect to the plane of the membrane.  We see distinct
  peaks in the distribution of these angles (Figure~\ref{fig:tail_packing}).
